---
title: "Inference Simulations: solutions"
subtitle: "PSYC 201A, Simulation Lab 2"
author: "Bria Long"
date: today
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Tips before getting started

This is a document made to accompany some simulations for the Inference workshop in Psych 201a. The goal of this document is to continue learning in R/tidyverse but also to gain hands on experience simulating and manipulating data.\
\
Open this file by double-clicking `inference_simulations_lab_2.Rproj` first. The only package it needs is the tidyverse.

To understand what a function does, type `?function_name` in the Console (e.g., `?rnorm`).

# Setup

```{r}
set.seed(123) # good practice to set a random seed, or else different runs get you different results
```

## Import the functions & data that we need

```{r}
#| eval: false
install.packages("tidyverse")
```

```{r}
library(tidyverse)
```

# Part 1: Simulating experimental data

## Define the simulation function
This makes "tea data", a tibble (dataframe) where there are a certain number of people in each condition (default = 48, i.e., n_total, with n_total/2 in each half)

The averages of the two conditions are separated by a known effect ("delta") with some variance ("sigma"). You can change these around since we're simulating data!

```{r}
make_tea_data <- function(n_total = 48,
                          sigma = 1.25,
                          delta = 1.5) {
  n_half <- n_total / 2
  tibble(condition = c(rep("milk first", n_half), rep("tea first", n_half)),
         rating = c(round(
           rnorm(n_half, mean = 3.5 + delta, sd = sigma)
         ),
         round(rnorm(
           n_half, mean = 3.5, sd = sigma
         )))) |>
    mutate(rating = if_else(rating > 10, 10, rating),
           # truncate if greater than max/min of rating scale
           rating = if_else(rating < 1, 1, rating))
}
```

## Make data frames where we have small or larger samples of tea data for ONE experiment

```{r}
# here's, we're calling our custom function, and specifying different inputs than the defaults (which are inside the parenthese up above)
tea_data <- make_tea_data(n_total = 18, delta = .5)
```

```{r}
tea_data_highn <- make_tea_data(n_total = 48, delta = .5)
```

To do: OK, look at these data frames. How long are they, what are the column names? Look at them in your console and in the environment if you want.

To do: Write basic tidyverse code to calculate the mean of each condition (hint: use `group_by` and `summarize`)

```{r}
by_condition <- tea_data %>%
  group_by(condition) %>%
  summarize(mean_rating = mean(rating),
            num_participants = length(rating))

by_condition_highn <- tea_data_highn %>%
  group_by(condition) %>%
  summarize(mean_rating = mean(rating),
            num_participants = length(rating))
 
```

## In this first draw, was it significant with N=9 or N=24 per group?

OK, we can do t-tests already! I've done these within a pipe (whoops, I often use the old pipe operator because I have small hands)

To do: Run these and look at the outputs by posting `out_low_n` in the console

```{r}
out_low_n <- tea_data %>%
  t.test(rating ~ condition, data = .,  var.equal = TRUE)

out_high_n <- tea_data_highn %>%
  t.test(rating ~ condition, data = ., var.equal = TRUE) 
```

### Simulate 1000 experiments

To do: Run these code blocks

...where you have 18 participants per experiment with an average difference of 1.5 points in tea deliciousness on average

```{r}
samps <- tibble(sim = 1:1000) |> #
  mutate(data = map(sim, \(i) make_tea_data(n_total = 18, delta = 1.5))) |>  # simulate
  unnest(cols = data) # wrangle
```

...where you have 48 participants per experiment

```{r}
samps_highn <- tibble(sim = 1:1000) |> #
  mutate(data = map(sim, \(i) make_tea_data(n_total = 48, delta = 1.5))) |>  # simulate
  unnest(cols = data) # wrangle
```

### Summarize both of these simulations

To do: Run these code blocks Do you understand what each line is doing here? Sim = the simulated experiment number (arbitrary number from 1-1000)

```{r}
tea_data_summary <- samps |>
  group_by(sim, condition) |> # group by simulation #, and condition
  summarise(mean_rating = mean(rating)) |> # summarize across ratings
  group_by(sim) |> # now get difference
  summarise(delta = mean_rating[condition == "milk first"] -
              mean_rating[condition == "tea first"])
```

```{r}
tea_data_highn_summary <- samps_highn |>
  group_by(sim, condition) |> # group by simulation #, and condition
  summarise(mean_rating = mean(rating)) |> # summarize across ratings
  group_by(sim) |> # now get difference
  summarise(delta = mean_rating[condition == "milk first"] -
              mean_rating[condition == "tea first"])
```

### Plot difference for low-n

Let's make a plot to plot the differences in ratings across conditions To do: run these code blocks

```{r}
ggplot(data=tea_data_summary, aes(x=delta)) +
  geom_histogram(alpha=.8, bins=20) 
  
```

Or simply

```{r}
hist(tea_data_summary$delta)
```

### Plot difference for higher-n

To do: What's different about this distribution vs the one we just plotted?

```{r}
hist(tea_data_highn_summary$delta)
```

**Bonus** What happens if you run it again? Try varying the variance / mean of the effect when you change the "delta" and "sigma" values?

What happens if you vary the random seed?

# Part 2: Simulating the null distribution

(STOP HERE AND PAUSE FOR GROUP DISCUSSION)

Visualizing the null distribution. How can we actually visualize the NULL hypothesis and distribution? It's a little tricky, but we can rely on simulation to help us.

### Method 1: Simulation
First, do the same thing as you did before - simulating 1000 experiments -- but NOW simulate no differences between conditions. Remember, it's the null model because DELTA (i.e., differences in conditions) is ZERO. Keep these dataframes - we will use them later for estimating p-values


```{r}
null_model <- tibble(sim = 1:1000) |>
  mutate(data = map(sim, \(i) make_tea_data(n_total = 18, delta = 0))) |>
  unnest(cols = data)

null_model_summary <- null_model |>
  group_by(sim, condition) |>
  summarise(mean_rating = mean(rating)) |>
  group_by(sim) |>
  summarise(delta = mean_rating[condition == "milk first"] -
              mean_rating[condition == "tea first"])
  
```

Visualize what this null distribution actually looks like.

```{r}
ggplot(data=null_model_summary, aes(x=delta)) +
  geom_histogram(alpha=.8, bins=50) +
  theme_classic()
```

### Method 2: Shuffling

We can also calculate the distribution of the difference between conditions when we've shuffled the condition labels. This is referred to as the *empirical null hypothesis* (our H0, since we're comparing two conditions with a two-sample t-test)

### Let's shuffle the labels within each experiment

```{r}
tea_data_highn_shuffled <- samps_highn %>%
  group_by(sim) %>% # for each experiment
  mutate(condition_shuffled = sample(condition)) # shuffle the condition labels
```

To do: check your understanding -- what is "sim" here?

```{r}
tea_data_highn_shuffled_summary <- tea_data_highn_shuffled %>%
  group_by(condition_shuffled, sim) %>% 
  summarize(mean_rating = mean(rating),
            sd = sd(rating)) 

# overwrites this variable
tea_data_highn_shuffled_summary <- tea_data_highn_shuffled_summary %>%
  group_by(sim) |> 
  summarise(delta = mean_rating[condition_shuffled == "milk first"] -
              mean_rating[condition_shuffled == "tea first"]) # difference for each experimental draw
```

To do: what does this histogram look like and why?

```{r}
hist(tea_data_highn_shuffled_summary$delta) #what does this histogram look like?
```

### Visualizing what happens when we shuffle

OK, now see what happens to our raw data -- this is just from one simulation The color refers to the ORIGINAL label before we shuffled, but our condition difference is gone

The color refers to the ORIGINAL condition The x-axis refers to the SHUFFLED condition

```{r}
ggplot(data = tea_data_highn_shuffled %>% filter(sim==3),  # can change the actual simulation number here
       mapping = aes(x = condition_shuffled, y = rating))+
  geom_point(mapping = aes(color = condition), # color
             alpha=.8,
             position = position_jitter(height = .1,
                                        width = 0.1)) +
  stat_summary(fun.data = mean_se, fun.args = list(mult = 1.96), # mean +/- 1.96 SE: an approximate 95% CI
               geom = "linerange",
               linewidth = 1) +
  stat_summary(fun = "mean", # this calculates the average
               geom = "point",
               shape = 21,
               color = "black",
               fill = "white") + 
  scale_y_continuous(breaks = 0:10,
                     labels = 0:10,
                     limits = c(0, 10))
```

The idea is now that we can get a sampling distribution of the difference in the means between the two conditions (assuming that the null hypothesis were true), by randomly shuffling the labels and calculating the difference in means (and doing this many times).

What we get is a distribution of the differences we would expect, if there was no effect of condition.

First, let's calculate the actual difference in a simulated dataset where there was an effect

```{r}
difference_actual <- tea_data_highn |>  # in ONE experiment
  group_by(condition) |>
  summarize(mean_rating = mean(rating)) |>
  summarise(delta = mean_rating[condition == "milk first"] -
              mean_rating[condition == "tea first"]) |>
  pull(delta)
```

Plot the distribution of the differences across all of our shuffled iterations

And in red -- is the numerical value from when there was an effect (because we simulated it that way)
```{r}

ggplot(data = tea_data_highn_shuffled_summary, aes(x=delta)) +
  geom_histogram(aes(y = after_stat(density)),
                 color = "black",
                 fill = "lightblue",
                 binwidth = 0.05) + 
  stat_density(geom = "line",
               linewidth = 1.5,
               bw = 0.2) +
  geom_vline(xintercept = difference_actual, color = "red", linewidth = 2) +
  labs(x = "difference between means") 
  

```

And we can THEN calculate the p-value by using some basic data wrangling -- it is defined as the proportion of condition differences (treatment - control) that were as or more extreme than the one we observed.

If this feels tough, don't worry - we will go over these concepts more again on Friday!

```{r}
# two-sided: shuffled differences at least as far from zero as the observed one
tea_data_highn_shuffled_summary |>
  summarize(p_value = mean(abs(delta) >= abs(difference_actual)))
```

You can also see this if you plot the distributions of the null vs empirical simulations next to each other (blue = null, which is appropriately centered at zero. This should remind you of the cohen's d web app we looked at!)

```{r}
ggplot(data=tea_data_highn_shuffled_summary, aes(x=delta)) +
  geom_histogram(alpha=.4, bins=20, color='blue', fill='blue') +
  geom_histogram(alpha=.8, bins=20, data=tea_data_highn_summary)
```

# Part 3: Calculating confidence intervals

(STOP HERE AND PAUSE FOR GROUP DISCUSSION)

Honestly, I doubt you will get this far in lab -- but if you do, there are two extra pieces of material that are I think are foundational material.

## Confidence intervals

Here, I want to show you how you WOULD calculate a CI by hand. In practice, I never do this. But it's useful to see how it could be done.

### Get the data


Taking the high_n experiment for now

```{r}
# tea_dataset = tea_data
tea_dataset = tea_data_highn
```

```{r}
tea_ratings <- filter(tea_dataset, condition == "tea first")$rating
milk_ratings <- filter(tea_dataset, condition == "milk first")$rating

# could also do in a pipe like so, but then you have to grab the column below, as in tea_ratings$ratings; above is a vector
# tea_ratings <- tea_data_highn %>%
#   filter(condition=="tea first") %>%
#   select(rating)
```

## Calculate a CI on the effect (difference between conditions)
First we're going to calculate a CI by hand.


```{r}
n_tea <- length(tea_ratings)
n_milk <- length(milk_ratings)
sd_tea <- sd(tea_ratings)
sd_milk <- sd(milk_ratings)

tea_sd_pooled <- sqrt(((n_tea - 1) * sd_tea ^ 2 + (n_milk - 1) * sd_milk ^ 2) / 
                        (n_tea + n_milk - 2))

tea_se <- tea_sd_pooled * sqrt((1 / n_tea) + (1 / n_milk))


delta_hat <- mean(milk_ratings) - mean(tea_ratings)
tea_ci_lower <- delta_hat - tea_se * qnorm(0.975)
tea_ci_upper <- delta_hat + tea_se * qnorm(0.975)
```

## To get the 95% CI with the t-distribution

You need to get the appropriate t-statistic from the distribution, which incorporates information about the degrees of freedom

The t-distribution is more appropriate when you have smaller sample sizes and is what is used in t.tests

```{r}
num_observations = length(tea_dataset$rating)
df = num_observations-2 # for two sample t.test
tea_ci_lower_ttest <- delta_hat - tea_se * qt(0.975,df)
tea_ci_upper_ttest <- delta_hat + tea_se * qt(0.975,df)
```
The above code does this by hand - but it's to show you what this function is doing under the hood.


```{r}
# Now the calculated CIs match those in the t-test outputs!
t.test(milk_ratings, tea_ratings, var.equal = TRUE)
```

### Going to plot CIs for each condition

Also plot the SEs, and visualize how they're different

```{r}
confidence_level=.95 # you can change this
# this formula below gives the critical t-value (as opposed to simply taken from the normal distribution)
# qt(1 - (1 - confidence_level)/2, df = n - 1)


tea_data_highn_summary_cis <- tea_data_highn %>%
  group_by(condition) %>%
  summarize(cond_mean = mean(rating), cond_sd = sd(rating), n=length(rating)) %>%
  mutate(error = qt(1 - (1 - confidence_level)/2, df = n - 1)* (cond_sd/sqrt(n))) %>% # this calculates CIs WITHIN each condition
  mutate(ci_upper = cond_mean + error, ci_lower = cond_mean - error) %>%
  mutate(se_upper = cond_mean + cond_sd/sqrt(n), se_lower = cond_mean - cond_sd/sqrt(n))
```

Between subjects experiment -- lots of variability! I like to visualize the raw data as well as the mean and CIs

```{r}
ggplot(data = tea_data_highn, aes(x=condition, y=rating, col=condition))  +
  geom_jitter(width=.1, height=0, alpha=.3) + # visualizes all the raw data, with no variation in y-axis jitter
  theme_classic() +
  geom_pointrange(data = tea_data_highn_summary_cis, aes(x=condition, y = cond_mean, ymin = ci_lower, ymax = ci_upper)) +
  ylim(0,10) +
  ggtitle('Tea ratings across conditions with CIs')

```

```{r}
ggplot(data = tea_data_highn, aes(x=condition, y=rating, col=condition))  +
  geom_jitter(width=.1, height=0, alpha=.5) + # visualizes all the raw data, with no variation in y-axis jitter
  theme_classic() +
  geom_pointrange(data = tea_data_highn_summary_cis, aes(x=condition, y = cond_mean, ymin = se_lower, ymax = se_upper)) +
  ylim(0,10) +
  ggtitle('Tea ratings across condition with SE')

```

To do: What does each dot represnet? What does the range represent in each graph? What does the confidence interval indicate? What does the SE indicate?

To do: how does this change when you use the low-n experiment?

# Part 4: Simulating p-values across multiple experiments

### When there is no actual effect

To do: Comment what the lines of code are doing First for low-n experiments

```{r}
all_results=tibble() 

for (this_sim in 1:1000) {
  this_experiment = null_model %>%
    filter(sim==this_sim) 
  
  tea_ratings <- filter(this_experiment, condition == "tea first")$rating
  milk_ratings <- filter(this_experiment, condition == "milk first")$rating
  
  output = t.test(tea_ratings, milk_ratings, var.equal = TRUE)
  
  this_exp_output = tibble(pvalue = output$p.value)
  all_results = bind_rows(all_results, this_exp_output)
    
}
```

To do: Look at the distribution of p-values (hint: in all_results\$pvalue)

Make a histogram

What is the distribution of p-values when the null is true?

Calculate the proportion of p-values that are less than .05 What was our false positive rate?

```{r}
# you'll need all_results$pvalue
hist(all_results$pvalue)

# false positive rate: should be close to .05
mean(all_results$pvalue < .05)
```

### Now for an experiment when there is actually an effect

```{r}
all_results_high_n=tibble() 

for (this_sim in 1:500) {
  this_experiment = samps_highn %>%
    filter(sim==this_sim) 
  
  tea_ratings <- filter(this_experiment, condition == "tea first")$rating
  milk_ratings <- filter(this_experiment, condition == "milk first")$rating
  
  output = t.test(tea_ratings, milk_ratings, paired = FALSE, var.equal = TRUE)
  
  this_exp_output = tibble(pvalue = output$p.value)
  all_results_high_n = bind_rows(all_results_high_n, this_exp_output)
    
}
```

How often did we fail to reject the null hypothesis? When was our p-value greater than p=.05? What does the distribution of p-values look like?

```{r}
# you'll need all_results_high_n$pvalue
hist(all_results_high_n$pvalue)

# proportion of experiments that failed to reject the null (the miss rate; 1 minus this is power)
mean(all_results_high_n$pvalue >= .05)
```

# Additional Exercises

1.  Now go back (earlier in the code) and modify the DELTA in the simulation functions to be smaller so that there is only a small difference between groups. Is it still significant?

2.  Rewrite this code with the the smaller sample size simulations. What changes?
